---
title: Parallelizing simr::powercurve() in R
author: ~
date: ~
slug: parallelizing-simr-powercurve
categories:
  - R
  - research methods
tags:
  - s
  - power analysis
subtitle: ''
summary: 'Three strategies for reducing computation time when using simr::powercurve with Kenward-Roger or Satterthwaite methods: using high-performance computing, parallelizing different effects, and parallelizing different sample sizes within the breaks argument.'
authors: []
lastmod: '2021-07-23T16:46:54+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---


The `powercurve` function from the R package ['simr'](https://cran.r-project.org/web/packages/simr/simr.pdf) (Green & MacLeod, 2016) can incur very long running times when the method used for the calculation of *p* values is Kenward-Roger or Satterthwaite (see Luke, 2017). Here I suggest three ways for cutting down this time. 

1. Where possible, use a high-performance (or high-end) computing cluster. This removes the need to use personal computers for these long jobs.

2. In case you're using the `fixed()` parameter of the `powercurve` function, and calculating the power for different effects, run these at the same time ('in parallel') on different machines, rather than one after another. 

3. *Parallelize* the `breaks` argument. The `breaks` argument of the `powercurve` function allows the calculation of power for different levels of the grouping factor passed to `along`. Some grouping factors are *participant*, *trial* and *item*. The `breaks` argument sets the different sample sizes for which power will be calculated. Parallelizing `breaks` is done by running each number of levels in a separate function. When each has been run and saved, they are `c`ombined to allow the plotting. This procedure is  demonstrated below.


## Parallelizing `breaks`

Let's do a minimal example using a toy `lmer` model. A power curve will be created for the fixed effect of `x` along different sample sizes of the grouping factor `g`.

Notice that the six sections of the power curve below are serially arranged, one after another. In contrast, to enable parallel processing, each power curve would be placed in a single script, and they would all be run at the same time.

Although the power curves below run in a few minutes, the settings that are often used (e.g., a larger model; `fixed('x', 'sa')` instead of `fixed('x')`; `nsim = 500` instead of `nsim = 50`) take far longer. That is where parallel processing becomes useful.^[The number of simulations set by `nsim` should be larger (Brysbaert & Stevens, 2018; Green & MacLeod, 2016). In addition, the effect size for `x` should be adjusted to the value that best fits with the planned study (Kumle et al., 2021).]

```{r message = FALSE, warning = FALSE}

library(lme4)
library(simr)

# Toy model with data from 'simr' package
fm = lmer(y ~ x + (x | g), data = simdata)

# Extend sample size of `g`
fm_extended_g = extend(fm, along = 'g', n = 12)

# Parallelize `breaks` by running each number of levels in a separate function.

# 4 levels of g
pwcurve_4g = powerCurve(fm_extended_g, fixed('x'), along = 'g', breaks = 4, 
                        nsim = 50, seed = 123, 
                        # No progress bar
                        progress = FALSE)

# 6 levels of g
pwcurve_6g = powerCurve(fm_extended_g, fixed('x'), along = 'g', breaks = 6, 
                        nsim = 50, seed = 123, 
                        # No progress bar
                        progress = FALSE)

# 8 levels of g
pwcurve_8g = powerCurve(fm_extended_g, fixed('x'), along = 'g', breaks = 8, 
                        nsim = 50, seed = 123, 
                        # No progress bar
                        progress = FALSE)

# 10 levels of g
pwcurve_10g = powerCurve(fm_extended_g, fixed('x'), along = 'g', breaks = 10, 
                        nsim = 50, seed = 123, 
                        # No progress bar
                        progress = FALSE)

# 12 levels of g
pwcurve_12g = powerCurve(fm_extended_g, fixed('x'), along = 'g', breaks = 12, 
                        nsim = 50, seed = 123, 
                        # No progress bar
                        progress = FALSE)

```


Having saved each section of the power curve, we must now combine them to be able to plot them together (if you wish to automatise this procedure, consider [this function](https://github.com/pablobernabeu/Language-and-vision-in-conceptual-processing-Multilevel-analysis-and-statistical-power/blob/main/R_functions/combine_powercurve_chunks.R)).

```{r message = FALSE, warning = FALSE}

# Create a destination object using any of the power curves above.
all_pwcurve = pwcurve_4g

# Combine results
all_pwcurve$ps = c(pwcurve_4g$ps[1], pwcurve_6g$ps[1], pwcurve_8g$ps[1], 
                   pwcurve_10g$ps[1], pwcurve_12g$ps[1])

# Combine the different numbers of levels.
all_pwcurve$xval = c(pwcurve_4g$nlevels, pwcurve_6g$nlevels, pwcurve_8g$nlevels, 
                     pwcurve_10g$nlevels, pwcurve_12g$nlevels)

print(all_pwcurve)

plot(all_pwcurve, xlab = 'Levels of g')


# For reproducibility purposes
sessionInfo()

```


### Just the code

```

library(lme4)
library(simr)

# Toy model
fm = lmer(y ~ x + (x | g), data = simdata)

# Extend sample size of `g`
fm_extended_g = extend(fm, along = 'g', n = 12)

# Parallelize `breaks` by running each number of levels in a separate function.

# 4 levels of g
pwcurve_4g = powerCurve(fm_extended_g, fixed('x'), along = 'g', breaks = 4, 
                        nsim = 50, seed = 123, 
                        # No progress bar
                        progress = FALSE)

# 6 levels of g
pwcurve_6g = powerCurve(fm_extended_g, fixed('x'), along = 'g', breaks = 6, 
                        nsim = 50, seed = 123, 
                        # No progress bar
                        progress = FALSE)

# 8 levels of g
pwcurve_8g = powerCurve(fm_extended_g, fixed('x'), along = 'g', breaks = 8, 
                        nsim = 50, seed = 123, 
                        # No progress bar
                        progress = FALSE)

# 10 levels of g
pwcurve_10g = powerCurve(fm_extended_g, fixed('x'), along = 'g', breaks = 10, 
                        nsim = 50, seed = 123, 
                        # No progress bar
                        progress = FALSE)

# 12 levels of g
pwcurve_12g = powerCurve(fm_extended_g, fixed('x'), along = 'g', breaks = 12, 
                        nsim = 50, seed = 123, 
                        # No progress bar
                        progress = FALSE)


# Create a destination object using any of the power curves above.
all_pwcurve = pwcurve_4g

# Combine results
all_pwcurve$ps = c(pwcurve_4g$ps[1], pwcurve_6g$ps[1], pwcurve_8g$ps[1], 
                   pwcurve_10g$ps[1], pwcurve_12g$ps[1])

# Combine the different numbers of levels.
all_pwcurve$xval = c(pwcurve_4g$nlevels, pwcurve_6g$nlevels, pwcurve_8g$nlevels, 
                     pwcurve_10g$nlevels, pwcurve_12g$nlevels)


print(all_pwcurve)

plot(all_pwcurve, xlab = 'Levels of g')

```


## References

Brysbaert, M., & Stevens, M. (2018). Power analysis and effect size in mixed effects models: A tutorial. *Journal of Cognition, 1*(1), 9. http://doi.org/10.5334/joc.10

Green, P., & MacLeod, C. J. (2016). SIMR: An R package for power analysis of generalized linear mixed models by simulation. *Methods in Ecology and Evolution 7*(4), 493–498, https://doi.org/10.1111/2041-210X.12504

Kumle, L., Vo, M. L. H., & Draschkow, D. (2021). Estimating power in (generalized) linear mixed models: An open introduction and tutorial in R. *Behavior Research Methods*, 1--16. <https://doi.org/10.3758/s13428-021-01546-0>

Luke, S. G. (2017). Evaluating significance in linear mixed-effects models in R. *Behavior Research Methods, 49*(4), 1494--1502. <https://doi.org/10.3758/s13428-016-0809-y>
